### 2014-07-18 Thesis - 2/Paper/paper.tex
H5 + paper

In the end many choices can be made, but how does it reflect to the behaviour in practice? This paper will present result from experiments with HBase and MongoDB compared to the CAP theorem. The paper is organised as follows. Sect\todo{Do this}

\section{The CAP Theorem\cite{brewer2000towards}\cite{brewer2012cap}}
The CAP Theorem was introduced by E. Brewer \cite{brewer2000towards}  in 2000 and discusses 3 properties of which each network shared-data system can guarantee at most 2: (definitions based on \cite{brewer2012cap})
\begin{itemize}
\item (Strict)\textbf{C}onsistency: The system acts like there is only single storage 
\item High \textbf{a}vailability: The system is available (for updates)
\item \textbf{P}artition tolerance: A split in the network let the different partitions still act as a single system. 
\end{itemize}
Designers used this model to explain their design decisions, others used it to compare different system but sometimes it was misused. As E. Brewer explains 12 year after the launch, the "2 of 3" can be misleading.

@@

\section{Overview of HBase and MongoDB}
In this article, 2 CP systems will be discussed more in detail regarding their choices to forfeit availability and the influence on their behaviour in practice. The systems are HBase and MongoDB, an architectural overview will be given in this section.

@@

When a server goes down, he can release the lease in case there is a graceful shut down (the HBase server is notified) and another server can get the lease immediately. In other cases, a new lease can only be given after the decay of the old lease, if the server comes back on-line in meantime, he will still be responsible.

@@

The state of the different members of a ReplicaSet is maintained by a heartbeat system: a server is marked as off-line if no beat has been received for 10 seconds. In case the primary goes off-line, election will be started to re-elect a new one. In other words, a primary has a lease of 10 seconds. This value of 10 seconds is non-configurable.

@@

\section{Test method}
To test the behaviour of database systems towards consistency and availability, the Yahou Cloud Serving Benchmarking (YCSB \cite{cooper2010benchmarking}) has been extended with event support for availability and reader-writers for consistency.

@@

In these tests were 3 kinds of tests executed of which each one took 900 second. The first action takes place at 300 seconds, the second at 600 seconds.
\begin{itemize}
\item Graceful shut down of a data service and restart of the service
\item Hard kill of the data service and a restart of the service
\item Blocking of all network traffic from and to a server and the allowance of all network traffic
\end{itemize}

@@

These tests are executed on each of the data nodes of HBase and MongoDB.

\section{Results}\label{sec:result}
(50 threads, 600 queries/second) and MongoDB (15 threads, 200 queries/second). These values were based 
\section{Related work}

\section{Conclusion}
%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix

>>>>

New online services, more online users means more data, data and load a single server can't handle. More and more database systems are distributed, for higher availability in case of an unexpected crash but also for horizontal distribution: the data of a single database is spread over different systems. These new services have also different requirements, an update doesn't need to be visible for all users immediately, it can take time and the concept eventual consistency was there.

Over the past years, many new systems have been build on this wave of changes, and they are categorized under NoSQL. But some applications have a lot of data, the need to have the consistent data and being able to work in a highly distributed environment, called CP systems in the CAP Theorem. Two examples that offer these guarantees are HBase and MongoDB. They greatly differ in supported queries on their system, but what happens if you only use the bare essentials and compare their behaviour in a distributed environment going from expected shut downs of instances towards crashes and network partitions.

In this article, a comparison of both this systems on these 3 behaviours will be made. In chapter \ref{sec:CAPTheorem} a brief overview of the CAP Theorem is given, chapter \ref{sec:hbaseandmongodb} discusses HBase and MongoDB on paper. Chapter \ref{sec:testmethod} gives an overview of the used test method and chapter \ref{sec:result} presents the results. To end in chapter \ref{sec:futurework} future work is presented, an overview of related work is given in chapter \ref{sec:related work} and a conclusion is made in chapter \ref{sec:conclusion}.

\section{The CAP Theorem\cite{brewer2000towards}\cite{brewer2012cap}}\label{sec:CAPTheorem}
The CAP Theorem was introduced by E. Brewer \cite{brewer2000towards}  in 2000 and discusses 3 properties of which each network shared-data system can guarantee at most 2: (definitions based on \cite{brewer2012cap})
\begin{itemize}
\item (Strict)\textbf{C}onsistency: The system acts like there is only single storage 
\item High \textbf{a}vailability: The system is available (for updates)
\item \textbf{P}artition tolerance: A split in the network let the different partitions still act as a single system. 
\end{itemize}
Designers used this model to explain their design decisions, others used it to compare different system but sometimes it was misused. As E. Brewer explains 12 year after the launch, the "2 of 3" can be misleading.

@@

\section{Overview of HBase and MongoDB}\label{sec:hbaseandmongodb}
In this article, 2 CP systems will be discussed more in detail regarding their choices to forfeit availability and the influence on their behaviour in practice. The systems are HBase and MongoDB, an architectural overview will be given in this section.

@@

When a server goes down, he can release the lease in case there is a graceful shut down (the HBase server is notified) and another server can get the lease immediately. In other cases, a new lease can only be given after the decay of the old lease, if the server comes back online in meantime, he will still be responsible.

@@

The state of the different members of a ReplicaSet is maintained by a heartbeat system: a server is marked as offline if no beat has been received for 10 seconds. In case the primary goes offline, election will be started to re-elect a new one. In other words, a primary has a lease of 10 seconds. This value of 10 seconds is non-configurable.

@@

\section{Test method}\label{sec:testmethod}
To test the behaviour of database systems towards consistency and availability, the Yahou Cloud Serving Benchmarking (YCSB \cite{cooper2010benchmarking}) has been extended with event support for availability and reader-writers for consistency.

@@

Before the test are 300 000 records stored in the database to enable the sharding in all database softwares. In these tests were 3 kinds of tests executed of which each one took 900 second. The first action takes place at 300 seconds, the second at 600 seconds.
\begin{itemize}
\item Graceful shut down of a data service and restart of the service
\item Hard kill of the data service and a restart of the service
\item Blocking of all network traffic from and to a server and the allowance of all network traffic
\end{itemize}

@@

These tests are executed on each of the data nodes of HBase and MongoDB, caching and buffering on the client side are disabled in both tests.

\subsection{Consistency support}
To implement consistency support, the YCSB software is extended with an extra workload. A graphical representation of an example workload is shown in figure. 
\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{../img/Consistency-test-period}
\caption{Example of consistency support for one write period with 1 writer and 4 readers}
\end{figure}
These workloads exists out of 1 writer and a user-defined amount of readers.  Each writer will inserts or updates a value in the database with a user-defined period between them. The readers will read the record till they read the last written data element, each reader reads in a period defined by the user. The different readers are scheduled uniformly between the reading period.

All this data is logged and gives possibilities to analyse when a record is visible for different users, compared to the time the queries where started or ended.

Before the test are 30 000 records stored in the database and each tests takes 500 seconds and results start to be gathered after 30 seconds. In the tests is chosen to write every 0.5s and read with 5 readers every 10ms.

\section{Results}\label{sec:result}
To execute the tests, both systems were deployed on a virtual platform of OpenStack. Each instance has 2 CPUs, 4GB RAM and 50GB disc space. The machines are connected with a gigabit Ethernet and an average ping takes 0.4ms ($\sigma = 0.2$ on 10 000 ping's).

HBase is configured with 7 instances, of which 3 for management (1 HMaster, Hadoop namenode and Zookeeper, and 2 extra Zookeeper instances) and 4 for data storage (each has a HRegionServer and Hadoop datanode).

MongoDB is configured with 6 instance, grouped a 3 for a ReplicaSet. There was a single configuration server and 3 instances had an access server.

YCSB was deployed on a single instance and used to calibrate the systems to have a basic workload. An individual record has 10 fields which each field a size of 100 bytes. The workload existed out of 20\% inserts and updates, 40\% selects and 20\% scans of an uniform spread between 1 and 100. The requests are spread according to Zipfian\footnote{Some record are popular, others rare to be used.}  The basic load for HBase is an average of 600 queries/second spread over 50 threads, for MongoDB there are 15 threads with a total of 200 queries/second.

\subsection{Availability}
When reading and writing in the default setting of MongoDB, both MongoDB and HBase will read and write from a leader for a set of data. Both have a lease period for the leader, which can't be set for MongoDB (10 seconds), but can be configured for HBase (default 180 seconds).

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{../img/Observaties/HBase/Stop-07-14-Soft-3-1.png}
\caption{Example of gracefullyl shutting down a HBase node. The requests block for 90 seconds }
\end{figure}

In case of a stop of a HBase server in this configuration, the queries will halt till the lease for the region has been expired, this can take between 0 seconds and 180, depending on the moment of the action. Normally a release is possible in a graceful shut down, but as the pictures shows, this doesn't happen always.

For MongoDB, there is no difference between the graceful or hard stop of an instance; in case it was a secondary, no influences on the latency will be seen. In case a primary was stopped, will in both cases the data be temporarily unavailable.  In case of a network interruption, \todo{....}

\subsection{Consistency}
In consistency HBase and MongoDB have a different approach, more specific MongoDB has multiple configuration parameters for reading and writing while in H than HBase.

\section{Future work}\label{sec:futurework}

\section{Related work}\label{sec:related work}
The research towards database systems and their consistency guaranties is rare to have a measured approach. In recent paper (February 2014), Golab et al. states that there is only a limited amount of research done towards eventual consistency \cite{golab2014eventually}. They present a new view on consistency, were already some research is done towards active analyse (how long before the data is replicated to all nodes), the amount of passive analyse is limited (what do the users see). Compared to the consistency results from this paper, both are discussed: as well the delays before the data is present everywhere but also on the acting of the specific systems, in example the behaviour of HBase.

Another extension of YCSB called YCSB++\cite{patil2011ycsb++}, provides more logging information on all systems in the first place, but they also test the consistency of HBase in regards of the client caches. The reasoning for this is that it is the standard run configuration of HBase, but the submitting of the client cache towards the system depends not only on the time, but also on the amount of traffic that is being submitted. In the study they compare different cache sizes and this shows already a difference in time. Furthermore, it is possible to disable this caching in case there are records in the need of this strict consistency.

For availability benchmarking, there was no research found on related databases. However, research from 2004 \cite{mauro2004system} discuss a way to let a standalone system recover and provide a starting benchmark for it.

\section{Conclusion}\label{sec:conclusion}\todo{}

%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix

### 2014-07-18 Thesis - 2/hoofdstukken/5.observaties.tex
H5 + paper

Het aantal gebruikers wordt zo gekozen dat het totale aantal queries zakt of voor een sterke groei in vertraging zorgt, dit zorgt voor de gegevens in tabel \ref{table:calibratie-gebruikers-resultaat}. \todo{Check data in tabel}

\begin{table}[h!]
	\centering
	\begin{tabular}{l| l }
		\textbf{DBMS} & Aantal gebruikers \\
		\hline
		HBase & 40 \\
		MongoDB & 15\\
		Pgpool-II & 30\\
	\end{tabular}
	\caption{Calibratie: Aantal gebruikers per test voor de verschillende DBMS's}
	\label{table:calibratie-gebruikers-resultaat}
\end{table}

\begin{figure}[h!] 
\centering
	\subfigure[MongoDB]{\label{fig:calibratie-gebruikers-mongodb} \includegraphics[width=0.8\textwidth]{img/Observaties/threads-MongoDB}}
	\subfigure[Pgpool-II]{\label{fig:calibratie-gebruikers-pgpool-ii} \includegraphics[width=0.80\textwidth]{img/Observaties/threads-postgresql}}
	\subfigure[HBase]{\label{fig:calibratie-gebruikers-hbase} \includegraphics[width=0.8\textwidth]{img/Observaties/threads-HBase}}
	\caption{Calibratie: Overzicht van het aantal requests tot de gemiddelde vertraging voor verschillend aantal gebruikers. Elk datapunt stelt een verschillend aantal gebruikers voor met het aantal rechtsboven het punt. }
	\label{fig:calibratie-gebruikers-resultaat}
\end{figure}

@@

Met beide figuren samen, kan een matige belasting gekozen. Een matige belasting is een belasting waarbij de onderste figuur de waarde 1 zo dicht mogelijk benaderd en de vertraging nog niet te veel is gestegen t.o.v. van een lage belasting. De gekozen waarde zijn te vinden in tabel \ref{table:calibratie-queriesperseconde-resultaat}.  \todo{Check data in tabel}

\begin{table}[h!]
	\centering
	\begin{tabular}{l| l }
		\textbf{DBMS} & Aantal requests per seconde \\
		\hline
		HBase & 200 \\
		MongoDB & 200\\
		Pgpool-II & 100\\
	\end{tabular}
	\caption{Calibratie: Aantal queries per seconde per test bij een matige belasting voor de verschillende DBMS's.}
	\label{table:calibratie-queriesperseconde-resultaat}
\end{table}

\begin{figure}[h!] 
	\centering
	\subfigure{\label{fig:calibratie-queriesperseconde-hbase-1} \includegraphics[width=0.8\textwidth]{img/Observaties/loadbalance-db-HBase}}
	\subfigure{\label{fig:calibratie-queriesperseconde-hbase-2} \includegraphics[width=0.8\textwidth]{img/Observaties/loadbalance-realthroughput-db-HBase}}
	\caption{Calibratie: Overzicht van de vertraging t.o.v. het theoretisch aantal aanvragen met een vergelijking hoeveel werkelijke aanvragen er waren voor HBase. }
	\label{fig:calibratie-queriesperseconde-hbase}
\end{figure}

\begin{figure}[h!] 
	\centering
	\subfigure{\label{fig:calibratie-queriesperseconde-mongodb-1} \includegraphics[width=0.8\textwidth]{img/Observaties/loadbalance-db-MongoDB}}
	\subfigure{\label{fig:calibratie-queriesperseconde-mongodb-2} \includegraphics[width=0.8\textwidth]{img/Observaties/loadbalance-realthroughput-db-MongoDB}}
	\caption{Calibratie: Overzicht van de vertraging t.o.v. het theoretisch aantal aanvragen met een vergelijking hoeveel werkelijke aanvragen er waren voor MongoDB. }
	\label{fig:calibratie-queriesperseconde-mongodb}
\end{figure}

\begin{figure}[h!] 
	\centering
	\subfigure{\label{fig:calibratie-queriesperseconde-pgpool-ii-1} \includegraphics[width=0.8\textwidth]{img/Observaties/loadbalance-db-PostgreSQL}}
	\subfigure{\label{fig:calibratie-queriesperseconde-pgpool-ii-2} \includegraphics[width=0.8\textwidth]{img/Observaties/loadbalance-realthroughput-db-PostgreSQL}}
	\caption{Calibratie: Overzicht van de vertraging t.o.v. het theoretisch aantal aanvragen met een vergelijking hoeveel werkelijke aanvragen er waren voor Pgpool-II. }
	\label{fig:calibratie-queriesperseconde-pgpool-ii}
\end{figure}

\section{Beschikbaarheidstest}
Bij de beschikbaarheidstesten kunnen de gegevens op 3 manieren voorgesteld worden: de vertraging per query over de hele test, de vertraging tijdens het stoppen en starten van systemen of een vergelijking van de vertraging voor het stoppen, na het herstarten en tussen het stoppen en starten.

@@

\paragraph{HBase} \todo{}

\paragraph{MongoDB} \todo{}

\paragraph{PostgreSQL} \todo{}

\section{Consistentie test}

>>>>

Het aantal gebruikers wordt zo gekozen dat het totale aantal queries zakt of voor een sterke groei in vertraging zorgt, dit zorgt voor de gegevens in tabel \ref{table:calibratie-gebruikers-resultaat}. Bij MongoDB is er voor een lage waarde van 15 gebruikers gekozen, in plaats van 50, de reden hiervoor is dat de variatie in de vertraging groter wordt bij meer gebruikers, wat de testen moeilijker maakt.

\begin{table}[h!]
	\centering
	\begin{tabular}{l| l }
		\textbf{DBMS} & Aantal gebruikers \\
		\hline
		HBase & 50 \\
		MongoDB & 15\\
		Pgpool-II & 30\\
	\end{tabular}
	\caption{Calibratie: Aantal gebruikers per test voor de verschillende DBMS's}
	\label{table:calibratie-gebruikers-resultaat}
\end{table}

\begin{figure}[h!] 
\centering
	\subfigure[MongoDB]{\label{fig:calibratie-gebruikers-mongodb} \includegraphics[width=0.4\textwidth]{img/Observaties/threads-MongoDB}}
	\subfigure[Pgpool-II]{\label{fig:calibratie-gebruikers-pgpool-ii} \includegraphics[width=0.4\textwidth]{img/Observaties/threads-postgresql}}
	\subfigure[HBase]{\label{fig:calibratie-gebruikers-hbase} \includegraphics[width=0.4\textwidth]{img/Observaties/threads-HBase}}
	\caption{Calibratie: Overzicht van het aantal queries tot de gemiddelde vertraging voor verschillend aantal gebruikers. Elk datapunt stelt een verschillend aantal gebruikers voor met het aantal rechtsboven het punt. }
	\label{fig:calibratie-gebruikers-resultaat}
\end{figure}

@@

Met beide figuren samen, kan een matige belasting gekozen. Een matige belasting is een belasting waarbij de onderste figuur de waarde 1 zo dicht mogelijk benaderd en de vertraging nog niet te veel is gestegen t.o.v. van een lage belasting. De gekozen waarde zijn te vinden in tabel \ref{table:calibratie-queriesperseconde-resultaat}.

\begin{table}[h!]
	\centering
	\begin{tabular}{l| l }
		\textbf{DBMS} & Aantal requests per seconde \\
		\hline
		HBase & 600 \\
		MongoDB & 200\\
		Pgpool-II & 100\\
	\end{tabular}
	\caption{Calibratie: Aantal queries per seconde per test bij een matige belasting voor de verschillende DBMS's.}
	\label{table:calibratie-queriesperseconde-resultaat}
\end{table}

\begin{figure}[h!] 
	\centering
	\subfigure{\label{fig:calibratie-queriesperseconde-hbase-1} \includegraphics[width=\textwidth]{img/Observaties/loadbalance-db-HBase}}
	\subfigure{\label{fig:calibratie-queriesperseconde-hbase-2} \includegraphics[width=\textwidth]{img/Observaties/loadbalance-realthroughput-db-HBase}}
	\caption{Calibratie: Overzicht van de vertraging t.o.v. het theoretisch aantal aanvragen met een vergelijking hoeveel werkelijke aanvragen er waren voor HBase. }
	\label{fig:calibratie-queriesperseconde-hbase}
\end{figure}

\begin{figure}[h!] 
	\centering
	\subfigure{\label{fig:calibratie-queriesperseconde-mongodb-1} \includegraphics[width=\textwidth]{img/Observaties/loadbalance-db-MongoDB}}
	\subfigure{\label{fig:calibratie-queriesperseconde-mongodb-2} \includegraphics[width=\textwidth]{img/Observaties/loadbalance-realthroughput-db-MongoDB}}
	\caption{Calibratie: Overzicht van de vertraging t.o.v. het theoretisch aantal aanvragen met een vergelijking hoeveel werkelijke aanvragen er waren voor MongoDB. }
	\label{fig:calibratie-queriesperseconde-mongodb}
\end{figure}

\begin{figure}[h!] 
	\centering
	\subfigure{\label{fig:calibratie-queriesperseconde-pgpool-ii-1} \includegraphics[width=\textwidth]{img/Observaties/loadbalance-db-PostgreSQL}}
	\subfigure{\label{fig:calibratie-queriesperseconde-pgpool-ii-2} \includegraphics[width=\textwidth]{img/Observaties/loadbalance-realthroughput-db-PostgreSQL}}
	\caption{Calibratie: Overzicht van de vertraging t.o.v. het theoretisch aantal aanvragen met een vergelijking hoeveel werkelijke aanvragen er waren voor Pgpool-II. }
	\label{fig:calibratie-queriesperseconde-pgpool-ii}
\end{figure}

\section{Beschikbaarheidstest}
Bij de beschikbaarheidstesten kunnen de gegevens op verschillende manieren voorgesteld worden: de vertraging per query over de hele test, de vertraging tijdens het stoppen en starten van systemen of een vergelijking van de vertraging voor het stoppen (150-250s), na het herstarten (700-800s) en tussen het stoppen en starten(400-500s).

@@

Een punt op de grafiek stelt de gemiddelde vertraging van 1 seconde voor, de lijn het gemiddelde over 10 seconden.

\paragraph{HBase}
Bij HBase zijn er verschillende reacties op het stopzetten van een node. Bij het zacht stoppen van een instantie, is er een onderbreking van gemiddeld 20 seconde in de testen. Daarna kunnen er nog verhogingen in de queries af en toe optreden. Zie figuur \ref{fig:beschikbaar-hbase-soft}.

Bij de netwerk onderbreking is er in de testen een onderbreking van ongeveer 100 seconden. Daarna is het terug stabiel. Zie figuur \ref{fig:beschikbaar-hbase-drop}

Bij een harde stop is er een combinatie van de netwerk onderbreking én is het af en toe zo dat de volledige periode geen queries mogelijk zijn. Zie figuur \ref{fig:beschikbaar-hbase-drop} en \ref{fig:beschikbaar-hbase-hard}

Tijdens de onderbreking, is er geen significante verandering in de vertraging van de uitgevoerde queries gemeten. 
\begin{figure}[ht!] 
	\centering
	\subfigure[Voorbeeld zacht stop]{\label{fig:beschikbaar-hbase-soft} \includegraphics[width=\textwidth]{img/Observaties/HBase/single-graph-2-1}}
	\subfigure[Voorbeeld netwerk onderbreking en harde stop]{\label{fig:beschikbaar-hbase-drop} \includegraphics[width=\textwidth]{img/Observaties/HBase/single-graph-2-drop-1}}
	\subfigure[Voorbeeld hard stop]{\label{fig:beschikbaar-hbase-hard} \includegraphics[width=\textwidth]{img/Observaties/HBase/single-graph-3-kill-1}}
	\caption{Beschikbaarheid: Verschillende voorbeeldreacties van HBase op beschikbaarheidstesten }
	\label{fig:beschikbaar-hbase-1}
\end{figure}

\paragraph{MongoDB} \todo{}
Ook bij MongoDB zijn er verschillende reacties op het stopzetten als men de queries onder standaard configuratie uitvoert. In het geval van zacht of hard stoppen is er geen verschil in de reactie, bij 2/3 van de keren is er geen verschil merkbaar, bij 1/3 van de keren is er tijdelijke verhoging van het aantal queries, een voorbeeld toont dat de scan operatie voor 2 seconden uitgesteld wordt. Figuur van het overzicht: \ref{fig:beschikbaar-mongodb-soft} met een zoom naar de stop \ref{fig:beschikbaar-mongodb-soft-zoom}.

Bij het onderbreken van het netwerk is er in bepaalde geen significante verandering, op andere momenten is een gedrag soortgelijk aan dat bij een zachte stop te merken. In andere gevallen is het zo dat er geen queries mogelijk zijn gedurende de volledige netwerk onderbreking. Zie figuur \ref{fig:beschikbaar-mongodb-drop}.

Tijdens de onderbreking, is er geen significante verandering in de vertraging van de uitgevoerde queries gemeten. 
\begin{figure}[ht!] 
	\centering
	\subfigure[Voorbeeld van een zacht stop, harde stop en netwerk onderbreking]{\label{fig:beschikbaar-mongodb-soft} \includegraphics[width=\textwidth]{img/Observaties/MongoDB/single-graph-6-1}}
	\subfigure[Zacht stop met inzoomen op het uitschakelen]{\label{fig:beschikbaar-mongodb-soft-zoom} \includegraphics[width=\textwidth]{img/Observaties/MongoDB/multiple-graph-interrupt-Node-6-Shut-down}}
	\subfigure[Voorbeeld netwerk onderbreking ]{\label{fig:beschikbaar-mongodb-drop} \includegraphics[width=\textwidth]{img/Observaties/MongoDB/single-graph-4-drop-1}}
	\caption{Beschikbaarheid: Verschillende voorbeeldreacties van MongoDB op beschikbaarheidstesten }
	\label{fig:beschikbaar-mongodb-1}
\end{figure}

\paragraph{Pgpool-II} Bij Pgpool-II is een gr\todo{}

In het geval van een zachte stop, is er tijdelijk een onderbreking van al de queries, de queries worden tijdelijk vertraagd met ongeveer 2 seconden, een voorbeeld bevindt zich in figuur \ref{fig:beschikbaar-pgpool-soft}.

Bij een netwerk onderbreking, zijn er enige tijd geen queries mogelijk en na 30 seconden was de onderbreking over in de testen.  Een voorbeeld bevindt zich in figuur \ref{fig:beschikbaar-pgpool-netwerk}.

Tijdens de onderbreking is er een verandering naar schrijf queries toe, deze nemen significant minder tijd in beslag, dit geldt niet voor lees queries. Het voorbeelden uit de testen bevinden zich in figuur \ref{fig:beschikbaar-pgpool-boxplot-read} en \ref{fig:beschikbaar-pgpool-boxplot-write} voor respectievelijk schrijf en lees queries. 
\begin{figure}[ht!] 
	\centering
	\subfigure[Voorbeeld van een zacht stop]{\label{fig:beschikbaar-pgpool-soft} \includegraphics[width=\textwidth]{img/Observaties/Pgpool/single-graph-1-2}}
	\subfigure[Voorbeeld van een harde stop of netwerk onderbreking]{\label{fig:beschikbaar-pgpool-netwerk} \includegraphics[width=\textwidth]{img/Observaties/Pgpool/single-graph-1-drop-1}}
	\subfigure[Boxplot met leesvertragingen]{\label{fig:beschikbaar-pgpool-boxplot-read} \includegraphics[width=0.45\textwidth]{img/Observaties/Pgpool/boxplot-graph-Node-1-READ}}
	\subfigure[Boxplot met schrijf vertragingen]{\label{fig:beschikbaar-pgpool-boxplot-write} \includegraphics[width=0.45\textwidth]{img/Observaties/Pgpool/boxplot-graph-Node-1-UPDATE}}
	\caption{Beschikbaarheid: Verschillende voorbeeldreacties van MongoDB op beschikbaarheidstesten }
	\label{fig:beschikbaar-pgpool}
\end{figure}

\section{Consistentie test}
Voor de consistentie testen worden er empirische verdelingsfuncties gebruikt. Dit zijn functies waarbij op de x-as een waarde verloop is en op de y-as het percentage van de waarden kleiner dan x staat aangegeven.

Voor de consistentie testen wordt er op de x-as de start- en/of stoptijdstippen van de verschillende soorten getoond. Het verschil tussen de y-waarde van de start- en stoptijdstippen geeft aan hoeveel queries er op dat moment uitgevoerd worden. De startmomenten van een lezer zijn de eerste keer dat deze de correcte data leest.

\paragraph{HBase}
Bij HBase is er geen verschil tussen het invoegen of aanpassen van data naar consistentie, dit zijn dezelfde queries. Daarnaast zijn er geen configuratie mogelijkheden voor het lezen of schrijven van data naast het in- of uitschakelen van de caches aan de gebruikerskant.

Figuur \ref{fig:consistentie-hbase-start} toont een overzicht van de verschillende starttijdstippen voor het lezen van consistente data. Figuur \ref{fig:consistentie-hbase-R1} toont de start- en eindtijdstippen voor lezer 2 naast deze voor de schrijver. De maximale waarde van de x-as is zo gekozen dat voor elke dataset minstens 99\% van de data getoond is.

\begin{figure}[ht!] 
	\centering
	\subfigure[Starttijdstippen van verschillende lezers voor lezen van consistente data]{\label{fig:consistentie-hbase-start} \includegraphics[width=.45\textwidth]{img/Observaties/HBase/ECDF-plot-Start-insertRawData-5}}
	\subfigure[De start- en stoptijdstippen voor lezer 2]{\label{fig:consistentie-hbase-R1} \includegraphics[width=0.45\textwidth]{img/Observaties/HBase/ECDF-plot-R-1-insertRawData-5}}

	\caption{Consistentie: Overzicht van HBase op de consistentie testen met een 99-percentiel.}
	\label{fig:consistentie-hbase}
\end{figure}

\paragraph{MongoDB} Bij MongoDB zijn er 5 soorten lees- en 5 soorten schrijfconfiguraties mogelijk. Na de testen bleek het dat secondarypreferred gelijk was aan secondary en is niet getoond in de grafieken om deze reden.

Voor elk van de 5 mogelijke schrijvacties, is een overzicht gegeven in figuren \ref{fig:consistentie-mongodb-all} en \ref{fig:consistentie-mongodb-R2}. In de eerste figuur, is de data van al de lezers gecombineerd, bij de tweede wordt er enkel naar lezer 2 gekeken. De maximale waarde van de x-as is zo gekozen dat voor elke dataset minstens 97\% en 99\% van de data getoond is voor respectievelijk figuur \ref{fig:consistentie-mongodb-all} en \ref{fig:consistentie-mongodb-R2}. Er zijn telkens de update queries getoond met een vergelijking tussen de insert en update bij majority. Ook uit de andere data blijkt dat deze niet significant verschillend zijn.

Een vergelijking van de duur van de verschillende leesoperaties kan gevonden worden in figuur \ref{fig:consistentie-mongodb-all-mongodb-write} waar er minstens 90\% van de data is getoond.

\begin{figure}[htb!] 
	\centering
	\includegraphics[width=.42\textwidth]{img/Observaties/MongoDB/ECDF-Compare-Write-insert-1}
	\caption{Consistentie: Overzicht van MongoDB's verschillende schrijfoperaties met een 90-percentiel. }
	\label{fig:consistentie-mongodb-all-mongodb-write}
\end{figure}

\begin{figure}[ht!] 
	\centering
	\subfigure[Normal Update]{\label{fig:consistentie-all-mongodb-normal} \includegraphics[width=.42\textwidth]{img/Observaties/MongoDB/ECDF-Reads-update-normal-all-2}}
	\subfigure[Safe Update]{\label{fig:consistentie-all-mongodb-safe} \includegraphics[width=.42\textwidth]{img/Observaties/MongoDB/ECDF-Reads-update-safe-all-2}}
	\subfigure[Fsync Safe Update]{\label{fig:consistentie-all-mongodb-fsync} \includegraphics[width=.42\textwidth]{img/Observaties/MongoDB/ECDF-Reads-update-fsync_safe-all-2}}
	\subfigure[Replica Safe Update]{\label{fig:consistentie-all-mongodb-replicasafe} \includegraphics[width=.42\textwidth]{img/Observaties/MongoDB/ECDF-Reads-update-replicas_safe-all-2}}
	\subfigure[Majority Update]{\label{fig:consistentie-mongodb-all-majority} \includegraphics[width=.42\textwidth]{img/Observaties/MongoDB/ECDF-Reads-update-majority-all-2}}
	\subfigure[Majority Insert]{\label{fig:consistentie-mongodb-all-majority-insert} \includegraphics[width=.42\textwidth]{img/Observaties/MongoDB/ECDF-Reads-insert-majority-all-2}}
	\caption{Consistentie: Overzicht van MongoDB op de consistentie testen voor alle lezers gecombineerd met een 97-percentiel }
	\label{fig:consistentie-mongodb-all}
\end{figure}

\begin{figure}[ht!] 
	\centering
	\subfigure[Normal Update]{\label{fig:consistentie-mongodb-R2-normal} \includegraphics[width=.42\textwidth]{img/Observaties/MongoDB/ECDF-Reads-update-normal-1-2}}
	\subfigure[Safe Update]{\label{fig:consistentie-mongodb-R2-safe} \includegraphics[width=.42\textwidth]{img/Observaties/MongoDB/ECDF-Reads-update-safe-1-2}}
	\subfigure[Fsync Safe Update]{\label{fig:consistentie-mongodb-R2-fsync} \includegraphics[width=.42\textwidth]{img/Observaties/MongoDB/ECDF-Reads-update-fsync_safe-1-2}}
	\subfigure[Replica Safe Update]{\label{fig:consistentie-mongodb-R2-replicasafe} \includegraphics[width=.42\textwidth]{img/Observaties/MongoDB/ECDF-Reads-update-replicas_safe-1-2}}
	\subfigure[Majority Update]{\label{fig:consistentie-mongodb-R2-majority} \includegraphics[width=.42\textwidth]{img/Observaties/MongoDB/ECDF-Reads-update-majority-1-2}}
	\subfigure[Majority Insert]{\label{fig:consistentie-mongodb-R2-majority-insert} \includegraphics[width=.42\textwidth]{img/Observaties/MongoDB/ECDF-Reads-insert-majority-1-2}}
	\caption{Consistentie: Overzicht van MongoDB op de consistentie testen voor lezer 2 met een 99-percentiel}
	\label{fig:consistentie-mongodb-R2}
\end{figure}

\section{Conclusie}
Uit de testen
